\MOC Server Pages:

### 2007-04-16 paper/paper.tex
svn-r25: paper updated -- as submitted to MetaOCaml workshop
darcs-hash:20070416223017-e5a07-ecd5a932f5585790b224cf5f4e386aa60cf4cb5f.gz

In their infancy (ca.~1993), web sites were little more than
clickable FTP sites with in-line images.  They enabled
distribution of static documents, but did little to enhance
collaboration or create a sense of community.  In contrast,
today's users expect most web sites to offer sophisticated
services and to permit endless personalization.  In 2004,
even a strictly personal web site is expected to incorporate
user-contributed content such as comments and ratings.

To provide such features, web sites contain custom computer
programs.  \citet{greenspun99panda}, in his insightful web
publishing guide book, describes the situation like this:

\begin{quote}
  You needn’t turn your site into a program just because the
  body of material that you are publishing is changing.
  Sites like Yahoo, for example, are sets of static files
  that are periodically generated by programs grinding
  through a dynamic database.  With this sort of
  arrangement, the site inevitably lags behind the database,
  but you can handle millions of hits a day without a major
  investment in computer hardware, custom software, or
  thought.\par If you want to make a collaborative site,
  however, then at least some of your web pages will have to
  be computer programs.  Pages that process user submissions
  have to add user-supplied data to your web server’s disk.
  Pages that display user submissions have to look through a
  database on your server before delivering the relevant
  contributions.~\cite[page 271]{greenspun99panda}
\end{quote}

In some notable ways, programming for the web is different
from programming in other domains.  Interaction design is
more of a challenge: in part because HTTP is stateless, but
also because users want to hit the `back' button and to set
bookmarks into the middle of
a carefully designed sequence of dialogs.%
\footnote{\citet{queinnec00influence} and
  \citet{graunke01programming} examine language-based
  approaches to solving this particular
  problem.} %
And unlike with shrink-wrapped applications, on the web it
is possible (though not always desirable) to ``go live''
immediately with a new fix or feature.

Nevertheless, there is at least one dictum of program design
that we cannot escape on the web: \emph{performance
  matters.}  As a web publisher, visitors are your bread and
butter.  But will your servers and scripts be ready for the
day that your site is featured on prime time television, or
on \texttt{slashdot.org}?  If tens of thousands of potential
users drop by to find a sluggish (or dead) server, most of
them will never return.

@@

Both of the preceding quotes point (informally) to the idea
of \emph{staging} the computation performed by a web
service.  Indeed, web publishing is an application area that
is naturally staged;\footnote{Nevertheless, it seems the
  literature on meta-programming has not yet acknowledged
  web services as a potential application area, although
  \citet{sheard01accomp} does
  mention \emph{mobile code.}} %
to wit:
\begin{enumerate}
\item Content (text, images, programs, etc.) created
  off-line is uploaded to the server---the \emph{publish}
  stage;

@@

During the \emph{serve} stage, we look up the user's topic
preferences, and ship out just the matching rows.  We delay
sorting the table until the \emph{display} stage.  The user
ought to be able to adjust the sort criteria without any
further communication with the server.  What about marking
past dates in red?  If this is also delayed until display,
then the code could be cached client-side for long periods
of time, yet still behave dynamically.

@@

This is in stark contrast to the status quo, where each
system targets one stage only.  The Website Meta Language
(WML) \cite{engelschall-wml} is an ``off-line HTML
generation toolkit'' designed for the \emph{publish} stage.
But many other programs (and countless ad-hoc scripts) spit
out HTML pages: think of \LaTeX2HTML.  Google
reports surprisingly many
programs\footnote{\url{http://google.com/search?q=gedcom+generate+html}}
for creating family tree web sites from genealogy database
files; these count as publish-stage tools.

The \emph{serve} stage is well-served by the ``server page''
languages, including JSP~\cite{sun-jsp},
ASP~\cite{microsoft-asp},
MSP~\cite{elsman03web}, and
PHP~\cite{bakken04php}.  The CGI protocol~\cite{w3c-cgi}
addresses the \emph{serve} stage, as do the embedded
interpreters (such as \texttt{mod\_perl} and
\texttt{mod\_python} for Apache) that exist to ameliorate some
of the overhead of CGI.
%%
There is, relatively, a paucity of languages that operate
client-side (\emph{display} stage), probably due to the
difficulty of securing an installed base of interpreters.
JavaScript~\cite{netscape-javascript} and Java applets are
notable exceptions.

Imagine implementing the conference calendar, as conceived
above, using currently deployed technology: A Perl script
outputs a PHP page which contains JavaScript!  Values are
passed from one stage to the next as strings, and the
programmer must manage all the quoting and persistence
issues by hand.

We present ``\MOC server pages,'' a new domain-specific
language for web applications programming.  It leverages the
staging annotations of \MOC~\cite{calcagno03meta,taha00metaml}
to provide safe and precise control over the each stage of the
computation.  We have implemented the system in two parts: our
translator transforms the server page language into a \MOC
module, which then can be incorporated into our multi-threaded
HTTP 1.1 server (also written in \MOC).  The scalability
gained by staging certain applications is stunning.

@@

\subsection{Translation to \MOC}
To see how all this works, let's consider how a \MOC server
page is translated into a proper \MOC program, to be executed
at publish time.  Since the program is executed before any
browser has requested the page, it cannot directly return or
output HTML.  Instead, it will construct and return a \cd{code}
object which is subsequently run on each request (serve stage).
Listing~\ref{lst:trans.meta} shows a sample \MOC server page,
and listing~\ref{lst:trans.ml} contains its translation.

@@

\mylst{trans.ml}{This is the translation of the
    page in figure~\ref{lst:trans.meta}.  Declarations have
    been lifted, and everything else appears within
    \cd{\Mbegincode brackets\Mendcode}}

Declaration blocks have been lifted to the top; any side effects
contained there are executed when the page is published.  The
constructed code begins with the \cd{{\KWfont let} page} line in
listing~\ref{lst:trans.ml}.

The \cd{a b c} represent publish-stage arguments (the names are
specified in the \cd{pragma}), whereas \cd{req} and \cd{puts} are
(fixed) serve-stage arguments.  \cd{req} encapsulates the HTTP request
details, including the headers and query arguments.  \cd{puts} is a
function, provided by the server, to 
transmit text across the network to the user's browser.  \cd{arg} is
then defined to look up an argument in the query string of the
request; its type is \cd{string\to string option}.

These are all \emph{serve} (second)
stage values.  It makes no sense to access them in the
\emph{publish} (first) stage, and indeed the \MOC type system
prevents this.

A few finer points to note: the translator discards newlines that
immediately follow code blocks (otherwise,
listing~\ref{lst:trans.ml} would be dotted with %
\cd{puts}~\verb+"\n"+ statements).  It automatically follows
code blocks with either a semi-colon or the {\KWfont in}
keyword, as required.

\subsection{Staged code blocks}
Now that we understand how the server page is assembled into a staged
program, the effects of adding \MOC staging operators to our pages
should be predictable.  Below is a reformulation of the permissions
example, using escape \Msplice{} and \cd{lift} to ensure that
\cd{stat} is called only once.%
\lgrindfile{perm2.meta}%
Still, the \cd{printf} conversion is performed at serve time, so maybe
it is best to pre-generate that as well:%
\lgrindfile{perm3.meta}%
Now, all that remains to execute at serve time is %
\cd{puts "0755"}.  Finally it is possible to splice in serve-stage
parameters at publish time, such as in this count-down example:%
\lgrindfile{count.meta}%
It removes the loop overhead by expanding to a sequence of 99
\cd{puts} statements.

@@

\subsection{Publish-stage arguments}
One last feature in the design of \MOC server pages is the ability
to take publish-stage arguments.  The parameter names are specified to
the translator in a \emph{pragma} declaration, which can appear
anywhere in the code:

(EXPLAIN THIS INSTEAD USING TRANS FIGURE)

Now the code can be instantiated with early parameters in countless
ways.  For example, requesting the URI \id{/longcount} could print a
count-down from 99, while \id{/shortcount} starts at 9 instead.  Both
are generated from the same server page.  Mapping from URIs to
instantiated \MOC code is, for now, an implementation detail
(see section~\ref{sec:impl}).

@@

\lgrindfile{navspec-sig.ml}
%%
The functions above may be invoked at any stage.  Naturally, if the
page title includes a dynamic argument, then \cd{preamble} will have
to be delayed until theh serve stage.

@@

@@

Judging from its prevalence in the literature, we are certain that
millions of grateful users would subscribe to an online service
capable of computing the \emph{power} function!  The screen shot in
figure~\ref{fig:power127} illustrates how it works.  The navigation
bar shows the exponents for which code has been
pre-generated.\footnote{In the current implementation, it is not
  possible for the user to request other exponents once the server is
  running.  See section~\ref{sec:impl} for an explanation.}  After
selecting the exponent, the user types the base into the form and
presses return.  The result is computed using OCaml's
arbitrary-precision \cd{Num} module.  The complete, documented script
appears in listing~\ref{lst:power.meta}.

The user's input is converted to a number relatively late in the
script.  If \cd{num\_of\_string} generates an exception (perhaps
because the user typed non-numeric text into the box), the navigation
bar and form will have already been output before the error message
appears.

@@

The \cd{list} is formed by testing filenames (in reverse alphabetical
order) against the compiled regular expression \cd{rc}, including only
those that match.  The file info record for each file has already
been prepared and appears in the code as a value.

@@

Currently, the \MOC implementation is not capable of reading
and writing \cd{code} values to a file.  This means that
generated server pages (after the publish stage) must already be
in the server memory when it starts.  This is an unfortunate
limitation, as it means we must restart the entire web server to
re-publish any particular page!

A generated server page has this type:

\lgrindfile{codeHandler.mli}

As the server starts, it constructs a dictionary (\cd{code
  StringMap.t}) mapping URIs to code.  This dictionary is
consulted on each HTTP request.  If the URI is not found there,
then we check the file system, in case it refers to a static
file.  If this fails, we respond with a 404 (resource not
found).

@@

Finally, we look at the performance of staged and unstaged directory
browsing; see figure~\ref{fig:browse}.  For this task, we created
directories containing a fixed number of files with random data.  The
average file size was 32k.  The \emph{x}~axis shows the number of
files in the directory.  Besides the staging factor, two
implementation languages are compared: we coded the same functionality
in PHP.  The `staged PHP' version must be run first from the command
line; this outputs another PHP script which is then run by the server.

@@

The degradation in performance of the staged PHP version is most
likely due to the fact that the stage two script must be \emph{parsed}
on each request; the size of the script is proportional to the number
of files in the directory.  With \MOC, the code size is also
proportional to the number of files, but the values are already in
memory.

@@

Web publishing is an important application domain that is
naturally staged.  Web programmers write staged programs every
day, but they do it the old-fashioned way: one script outputs
another as a string.  This is a tremendous opportunity for
multi-stage languages.

>>>>

Modern dynamic web sites support many features for user collaboration
and personalization.  To provide such services, web sites contain
custom computer programs, often written in one of a family of
programming languages that have grown up around (or been adapted for)
the web.

There is at least one dictum of program design that we cannot escape
on the web: \emph{performance matters.}  As a web publisher, visitors
are your bread and butter.  But will your servers and scripts be ready
for the day that your site is featured on prime time television, or on
\texttt{slashdot.org}?  If tens of thousands of potential users drop
by to find a sluggish (or dead) server, most of them will never
return~\cite{greenspun99panda}.

@@

This quote points (informally) to the idea of \emph{staging} the
computation performed by a web service.  Indeed, web publishing is an
application area that is naturally staged:\footnote{Nevertheless, it
  seems the literature on meta-programming has not yet acknowledged
  web services as a potential application area, although
  \citet{sheard01accomp} does
  mention \emph{mobile code.}} %
\begin{enumerate}
\item Content (text, images, programs, etc.) created
  off-line is uploaded to the server---the \emph{publish}
  stage;

@@

During the \emph{serve} stage, we look up the user's topic
preferences, and ship out just the matching rows.  If we delay
sorting the table until the \emph{display} stage, then the user
ought to be able to adjust the sort criteria without any
further communication with the server.  What about marking
past dates in red?  If this is also delayed until display,
then the code could be cached client-side for long periods
of time, yet still behave dynamically.

@@

This approach is in stark contrast to the status quo, where each
system targets one stage only.  One such system, the Website Meta
Language~\cite{engelschall-wml}, is an ``off-line HTML generation
toolkit'' designed for the \emph{publish} stage.  But many other
programs (and countless ad-hoc scripts) spit out HTML pages:
\LaTeX2HTML, for example.  Google reports surprisingly many
programs\footnote{\url{http://google.com/search?q=gedcom+generate+html}}
for creating family tree web sites from genealogy database files;
these also count as publish-stage tools.

The \emph{serve} stage is well-served by the ``server page''
languages, including JSP~\cite{sun-jsp},
ASP~\cite{microsoft-asp},
MSP~\cite{elsman03web}, and
PHP~\cite{bakken04php}.  The CGI protocol~\cite{w3c-cgi}
addresses the \emph{serve} stage, as do the embedded
interpreters (such as \texttt{mod\_perl} and
\texttt{mod\_python} for Apache) that exist to ameliorate some
of the overhead of CGI.

There is, relatively, a paucity of languages that operate
client-side (\emph{display} stage), probably due to the
difficulty of securing an installed base of interpreters.
JavaScript~\cite{netscape-javascript} and Java applets are
notable exceptions.

Imagine implementing the conference calendar, as conceived
above, using currently deployed technology: a Perl script
outputs a PHP page which contains JavaScript!  Values are
passed from one stage to the next as strings, and the
programmer must manage all the quoting and persistence
issues by hand.

We present ``\MOC server pages,'' a new domain-specific language for
web applications programming.  It leverages the staging annotations of
\MOC~\cite{calcagno03meta,taha00metaml} to provide safe and precise
control over the first two stages.  (We leave further consideration of
the display stage as future work.)  The system is implemented as two
components: a translator transforms the server page language into a
\MOC module, which then can be incorporated into our multi-threaded
HTTP/1.1 server (also written in \MOC).  The scalability gained by
staging certain applications is stunning.

@@

\subsection{Translation to \MOC}
To see how all this works, let's consider how a \MOC server
page is translated into a proper \MOC program, to be executed
at publish time.  Since the program is executed before any
browser has requested the page, it cannot directly return or
output HTML.  Instead, it will construct and return a \cd{code}
object which is subsequently run on each request (serve stage).
The third (display) stage proposed in section~\ref{sec:intro} is not
supported by this design.
Listing~\ref{lst:trans.meta} shows a sample \MOC server page,
and listing~\ref{lst:trans.ml} contains its translation.

@@

\mylst{trans.ml}{An automatic translation of the page in
  listing~\ref{lst:trans.meta}.  Declarations have been lifted, and
  everything else appears within \cd{\Mbegincode brackets\Mendcode}}

Declaration blocks have been lifted to the top; any side effects
contained there are executed when the page is published.  The
constructed code follows the line that begins with \cd{{\KWfont let} page}.

The \cd{a b c} represent publish-stage arguments (the names are
specified in the \cd{pragma}), whereas \cd{req} and \cd{puts} are
(fixed) serve-stage arguments.  \cd{req} encapsulates the HTTP request
details, including the headers and query arguments.  \cd{puts} is a
function, provided by the server, to transmit text across the network
to the user's browser.  \cd{arg} is then defined to look up an
argument in the query string of the request; its type is \cd{string\to
  string option}.
Since \cd{req} and \cd{puts} are serve-stage arguments, it is
incorrect to invoke them in the \emph{publish} (first) stage, and indeed
the \MOC type system prevents this.

A few finer points to note: the translator discards newlines that
immediately follow code blocks (otherwise,
listing~\ref{lst:trans.ml} would be dotted with %
\cd{puts}~\verb+"\n"+ statements).  It automatically appends a
semi-colon or the {\KWfont in} keyword to code blocks, as required (see,
for example, the lines containing \cd{statements},
\cd{more\_statements}, and \cd{{\KWfont let} x} in
listing~\ref{lst:trans.ml}).

\subsection{Staged code blocks}
Now that we understand how the server page is assembled into a staged
program, the effects of adding \MOC staging operators to our pages
should be predictable.  Below is a reformulation of the permissions
example, using escape \Msplice{} and \cd{lift} to ensure that
\cd{stat} is called only once.%
\lgrindfile{perm2.meta}%
Still, the \cd{printf} conversion is performed at serve time, so maybe
it is best to pre-generate that as well:%
\lgrindfile{perm3.meta}%
Now, all that remains to execute at serve time is %
\cd{puts "0755"}.  It is also possible to splice in serve-stage
parameters at publish time, such as in this count-down example:%
\lgrindfile{count.meta}%
It removes the loop overhead by expanding to a sequence of 99
\cd{puts} statements.

@@

\subsection{Publish-stage arguments}

\MOC server page support publish-stage arguments, as demonstrated by
the identifiers \cd{a b c} in listings~\ref{lst:trans.meta}
and~\ref{lst:trans.ml}.  The programmer specifies the pattern to be
used with a \cd{\Mquote{}pragma args ...\Munquote} declaration, which
may appear anywhere in the code.

With publish-stage arguments, the page may be instantiated in
countless ways.  Continuing the count-down example, we could map the
URI \id{/longcount} to a code block generated with an actual argument
of 99, while \id{/shortcount} refers to code with the actual
argument 9.  Both are generated from the same server page.

\lgrindfile{count2.meta}

Mapping from URIs to instantiated \MOC code is, for now, left as an
implementation detail (see section~\ref{sec:impl}).

@@

\lgrindfile{navspec-sig.ml}
%%
The functions above may be invoked at any stage.  Naturally, if the
page title includes a dynamic argument, then \cd{preamble} will have
to be delayed until the serve stage.

@@

Judging from its prevalence in the literature, we are certain that
millions of grateful users would subscribe to an online service
capable of computing the \emph{power} function!  The screen shot in
figure~\ref{fig:power127} illustrates how it works.  The navigation
bar shows the exponents for which code has been
pre-generated.\footnote{In the current implementation, it is not
  possible for the user to request other exponents once the server is
  running.  See section~\ref{sec:impl} for an explanation.}  After
selecting the exponent, the user types the base into the form and
presses return.  The result is computed using the arbitrary-precision
\cd{Num} module of the OCaml library.  The complete script appears in
listing~\ref{lst:power.meta}.

@@

The user's input is converted to a number relatively late in the
script; see the line marked with {(****)}.  If \cd{num\_of\_string}
generates an exception (perhaps because the user typed non-numeric
text into the box), the navigation bar and form will have already been
output before the error message appears.

@@

The \cd{list} is formed by testing filenames (in reverse alphabetical
order) against the compiled regular expression \cd{rc}, prepending only
those that match.  The file info record for each file has already
been prepared and appears in the code as a value.

@@

The implementation consists of two parts: a translator and a web
server.  The translator, implemented in Standard ML,\footnote{Mainly
  because the author has considerable experience with the ML-Yacc that
  comes with SML/NJ, and is reluctant to learn the OCaml version!}
transforms the server page syntax into plain \MOC, as illustrated in
listing~\ref{lst:trans.ml}.  It recognizes all the server page blocks
defined in section~\ref{sec:design} but does not completely parse the
\MOC code contained within them.  Instead, it recognizes just enough
of the keywords and delimiters to decide whether to add semi-colons or
`{\KWfont in}' after each block.  This approach makes the translator
fairly robust to minor changes in OCaml syntax.

The disadvantage of this method is that very few syntax errors (and
no type errors) are detected by the translator itself.  So, errors
reported by \MOC are displayed in terms of the translated program, not
the source program.

The server component is much more complex.  We implemented the
essential parts of the HTTP/1.1 specification as a multi-threaded
OCaml program.  Upon receiving a request for some URI, the server uses
a \emph{chain of responsibility}~\cite{gamma95design} to determine
how to handle it.  The chain is specified as a parameter to the
server.  Two primary handlers are provided: \cd{FileHandler} and
\cd{CodeHandler}.

The \cd{FileHandler} takes a file system path as a parameter, and then
interprets URIs as file names relative to that path.  If such a file
exists, it sends it verbatim to the client.  If not, it passes
the responsibility on to the next handler in the chain.

The \cd{CodeHandler} is provided with a dictionary to map URIs to code
values.  It looks up the URI in the dictionary, and if a match is
found it invokes that code.  For now, the map is hard-coded at build
time and the code values are already in memory when the server starts
accepting requests.  This is because \MOC is currently not capable of
reading and writing code values to a file.

Here are the types of the page code and the map, along with signatures
for the \cd{CodeHandler} (and two variants):

\lgrindfile{codeHandler.mli}
%%
\cd{page} is the type of the code that is constructed by each server
page; recall that the page takes two serve-time arguments: one
encapsulating the HTTP request, the other is the \cd{puts} function
for sending text back to the client.

The \cd{map} type maps a string (the URI) to a pair containing a
function to (re-)generate the page (by running the first stage) and a
ref cell where the page is cached.  No \cd{code} types show up here,
but they are used within the \cd{unit\to page} function.  This
function is implemented using the \Mrun{} (run) operator.  Passing
around code values themselves is difficult because they must remain
polymorphic to run them, and OCaml does not support rank-2
polymorphism.

@@

Finally, we look at the performance of staged and unstaged directory
browsing; see figure~\ref{fig:browse}.  Here, we created directories
containing fixed numbers of files with random data.  The average file
size was 32k.  The \emph{x}~axis shows the number of files in each
directory.

Besides the staging factor, two implementation languages are compared:
we implemented the same functionality in PHP.  The `staged PHP' version must
be run first from the command line; this outputs a PHP script
which is then run by the server.  With PHP we must, as noted in
section~\ref{sec:intro}, manage the quoting and persistence issues
by hand.  The staged PHP code contains gems like this:
\begin{verbatim}
<?= "<?\n" ?>
<?= "\$list = unserialize(\"".
    addcslashes(serialize($list),'"').
    "\");\n" ?>
<?= "?>\n" ?>
\end{verbatim}
where the serialize occurs in stage one and the unserialize in stage two.

@@

The degradation in performance of the staged PHP version is most
likely due to the fact that the stage two script must be \emph{parsed}
on each request; the size of the script is proportional to the number
of files in the directory.  With \MOC, the code size is also
proportional to the number of files, but the values are byte-compiled
(and already in memory).  At any rate, our goal is not to beat PHP on
performance, but rather to gain the performance benefit of staging
\emph{without} the awkwardness of staging in a language (like PHP) that
does not support it.

@@

Web publishing is an important application domain that is naturally
staged.  Web programmers write staged programs, but they do it the
old-fashioned way: one script outputs another as a string.  This is a
tremendous opportunity for multi-stage languages.
